Network Data Model and BerlinMOD Benchmark

### 2010-03-02 Algebras/Network/BerlinMODNetwork/paper/PaperBerlinMODAndNetworkDataModel.tex
Actualized to newest runtimes and query formulas.

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\var}[1]{\textsl{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\var{TRUE}}
\newcommand{\false}{\var{FALSE}}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document} 
\maketitle
\begin{abstract}
In the past several data models for the representation of spatial and spatio-temporal data objects have been developed. Among others we can categorise the data models into data models for objects moving freely in space and data models for moving objects that are constrained by a given network. For both categories several different data models have been developed. Two of them, one data model for objects moving freely in the 2D plane and one for network constrained moving objects have been implemented in \secondo{} DBMS. Both of them handling with the history data of moving objects. In this paper we present the comparison of the capabilities of the both data models using the \bmodb{}. The results show the advantages of specialised data models for network constrained objects. Therefore we propose a extension of the \bmodb{} with a additional query set BerlinMOD/Net including specialised challenges for network data models. This extensions enable us to compare the capabilities of different network data models with the \bmodb{}.
\end{abstract}
\section{Introduction}
In the past several data models for the representation of spatial and spatio-temporal data objects have been developed. Among others we can categorise them into data models for objects moving freely in space and data models for objects which movement is constrained by a network. For both categories several different data models have been presented like \cite{335426,chenzaniolosqlst} for spatio-temporal data objects moving freely in space and \cite{1146465,956692,VazWolfNetMod} for spatio-temporal data objects constrained by networks to name just a few.

For our experiments we choosed one example data model as representative for each of the both categories to compare the capabilities of the both different data models. The category of data models for spatio-temporal objects moving freely in space is represented by the data model presented in \cite{335426}. And the category of network data models is represented by the network data model presented in \cite{1146465}. The advantage of this choice is, that both data models are implemented in the \secondo{} DBMS and use the same temporal representation. So we can exclude that the use of different DBMS or temporal representations impact our results.

We used the recently introduced \bmodb{} \cite{BerlinMODVLDB} to compare the capabilities of the both data models. The \bmodb{} is best to our knowledge the first benchmark for complete spatio-temporal database systems and it is available in the \secondo{} DBMS. Furthermore the \bmodb{} data model is the data model of freely movement we use for our comparison. So we only have to transform the \bmodb{} spatial and spatio-temporal data types once in our network data model representation. This avoids sources of errors and relieves the control for correctness of the query results.

The results of our experiments show that it makes sense to use specialised network data models for network constrained objects, because the network data model outperforms the data model of free movement in the space in nearly all test cases. The total run time over all queries for the network data model is one-third of the total run time for the data model of free movement in the space. So the further developement of network data models seems to be useful. But network data models have their own challenges like network distance computing therefore we propose a extension of the \bmodb{} with a additional query set BerlinMOD/NET. This new query set should cover the specialised challenges of network data models. And enable us to compare the capabilities of different network data models also for the specialised network challenges with the \bmodb{}.

The rest of the paper is organised as follows. In section \ref{sec:relWork} we will first give short reminders of the underlying \secondo{} DBMS \ref{sec:secondo}, the \bmodb{} \ref{sec:bmodb} and the data models (\ref{sec:bmodbdatamod} and \ref{sec:netdatamod}) used in this paper. In section \ref{sec:bmodbNetDataMod} we present the setup for our experiments in \ref{sec:scenario} and describe our transfer of the \bmodb{} in the network data model representation in \ref{sec:Translation} before we show our experimental results in section \ref{sec:results}. In section \ref{sec:newqueries} we present our proposed extension of the \bmodb{} query sets. We conclude the paper with a summation of our results and aspects of future work in section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
As mentioned before in the past other data models for free movement in the space \cite{335426,chenzaniolosqlst} and for network constrained movement have been presented \cite{1146465,956692,VazWolfNetMod}. As well there are some more benchmarks \cite{COSTBenchmark, QueriesTheodoridis} and database systems for spatial and spatio-temporal datatypes \cite{HERMES,1054151}. But they don't provide a combination of different implemented and supported data types together with a existing benchmark like the actual \secondo{} DBMS. It is self-evident for us to use the \secondo{} DBMS in combination with the provided data types and benchmark to compare the capabilities of the both different representations of spatial and spatio-temporal data objects for network constrained objects.

In the next subsections we give short reminders of the used database system (\ref{sec:secondo}), the used benchmark (\ref{sec:bmodb}), and the used data models (\ref{sec:bmodbdatamod}, \ref{sec:netdatamod}). More detailed information for each of them can be taken from the original papers.
\subsection{Secondo}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in \cite{686903,1054151} provides a platform for implementing various kinds of data models defining a clean interface between the data model independent system frame and the content of the single data models. The data model independent components and mechanisms are separated from the data model dependent parts. Hence \secondo{} can be easy extended by implementing \secondo{} algebra modules to provide new data types and operations by the user. The user also may define additional viewers for the graphical user interface or write optimisation rules or cost functions to extend the optimiser. \secondo{} is free available in the web \cite{secondoweb} and comes with a number of already implemented spatial and spatio-temporal data types and operations including the spatio-temporal data model of free movement in the space \ref{sec:bmodbdatamod} and the network data model \ref{sec:netdatamod} (since version 2.8.4). Furthermore the \bmodb{} descripted in \ref{sec:bmodb} has been developed in the \secondo{} DBMS. For our experiments we used the \secondo{} version 2.9.1 with some minor bugfixes in the NetworkAlgebra and the TemporalNetAlgebra.
\subsection{BerlinMOD Benchmark}
\label{sec:bmodb}
The \bmodb{} was recently presented in \cite{BerlinMODVLDB} \nocite{BerlinMOD} and is implemented in the \secondo{} DBMS. It is available in the web \cite{berlinmodweb} and provides a well defined data-set and queries for the experimental evaluation of different moving object data representations. The \bmodb{} emphasises the development of complete systems and simplifies experimental repeatability pointing out the weakness and the potency of the benchmarked systems.

The data-sets of the \bmodb{} are created using the street map of Berlin \cite{bbike} and statistical data about the regions of Berlin \cite{bevberlin,berlinstadtatlas} as input relations. The created moving objects represent cars driving in the streets of Berlin. This makes it possible to use the data set of the \bmodb{} also for network constrained data models. Every moving object has a home node and a work node and every weekday there will be a trip from the home node to the work node in the morning and trip from the work node back to the home node in the afternoon. In the evening and at the weekend randomly chosen cars spend additional trips (one in the evening and up to six at the weekend) to different randomly chosen targets. The number of observed cars and the duration of the observation period can be influenced by the user of the \bmodb{} by setting the \var{scalefactor} to different values in the data generation script. For example at \var{scalefactor} 1.0 the data generator will create 2000 moving objects observed for 28 days. Each of them sending a GPS-signal every 2 seconds. This simulated signals are simplified so that time intervals when a car doesn't move or moves in the same direction at the same speed are merged into one unit. E.g. if the car holds 8 hours in front of the work node there will be only one entry in the cars history of movement with a time interval of 8 hours instead of 14.400 entries one for each GPS interval.

The \bmodb{} provides two different approaches to store the histories of moving objects. On the one hand the object-based approach (OBA) and on the other hand the trip based approach (TBA). In the OBA the complete history for each object is kept together. In the TBA exist several trips for each object. In the OBA there is only one relation (dataScar) containing one tuple for each object consisting of the spatio-temporal data of the object (journey), the licence, the type, and the model of the object. In the TBA we have two relations. One of them (dataMcar) contains the static data for each object like licence, type, and model together with an object identifier. The other relation (dataMtrip) contains for each object identifier several tuples each of them containing a single trip of the moving object, whereby breaks between the trips (e.g. the time the car holds in front of the office) are stored as single (stationary) trips.

Besides the moving objects the \bmodb{} provides sets of \var{QueryPoints}, \var{QueryRegions}, \var{QueryInstants}, \var{QueryPeriods}, and \var{QueryLicences}, each of them containing 100 objects (points, regions, time instants, time intervals, and licences) used in the benchmark queries.

For this data objects the \bmodb{} provides two sets of queries. One set addresses range queries (BerlinMOD/R) and the  other one nearest neighbour queries(BerlinMOD/NN). In this paper we focus on the range queries which are the main aspect of the \bmodb{} up to now. The query set BerlinMOD/R includes 17 queries selected of the set of possible combinations of the 5 aspects object identity (known / unknown), dimension (standard / spatial / temporal / spatio-temporal), query interval (point / range / unbounded), condition type (single object / object relations), and aggregation (with or without aggregation). We will present the 17 queries in more detail in section \ref{sec:queries} together with our network data model translations.
\subsection{BerlinMOD Data Model}
\label{sec:bmodbdatamod}
The data model used by the \bmodb{} is the same data model of freely moving in 2D space presented in \cite{594784,335426,352963}. All spatial positions are given in x,y-coordinates. A single spatial position is represented by the data type \dt{point}. A \dt{point} consists of a pair of \dt{real} values interpreted as x,y-coordinates in the 2D plane. Streets are represented by \dt{line} values. A \dt{line} value consists of a set of half segments representing the geometry of the line in the 2D plane. Each half segment consists of two \dt{point} values which are interpreted as start and end point of the half segment. Regions are represented by the data type \dt{region}. A \dt{region} consists of a set of half segments interpreted as outer (and inner) border of the region in the 2D plane.

@@

Some other data types of \secondo{} which are used in the \bmodb{} are shown in table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabular}{|l|p{9.5cm}|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type.\\
\hline
\dt{int} & Usual integer number.\\
\hline
\dt{real} & Usual real number.\\
\hline
\dt{instant} & A point in time.\\
\hline
\dt{periods} & A set of disjoint and not connected time intervals.\\
\hline
\dt{mbool} & A time dependent boolean value. The value within each \dt{ubool} will be constant \true{} or \false{} \\
\hline
\dt{mreal} & Time dependent real number. Each unit will be defined by a function of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabular}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\subsection{Network Data Model}
\label{sec:netdatamod}
The central idea of the network data model presented in \cite{1146465} is that every movement is constrained by a network and every position is given related to this network. Therefore the data type \dt{network} is the central data type in the network data model. All other data types of the network data model are related to a \dt{network} by the unique network identifier that is part of each \dt{network} object. The \dt{network} object contains all spatial information of the represented network in three main relations. The first relation contains the attributes of the routes (streets) like id, route curve, route length, and two boolean flags indicating if the route starts at the lexicographic smaller end point and if the lanes of the route are separated like on German Highways or not. The second relation contains all attributes of the junctions (street crossings) like the identifiers of the first and second route crossing in the junction, the distance of the junction from the start of the first respectively second route, tuple identifiers of the both routes in the routes relation, tuple identifiers of the sections connected by this junction in the sections relation, and a connectivity code telling us which lanes of the two routes are connected by the junction. The third main relation of a \dt{network} object is the sections relation containing the attributes of the sections (street parts between two junctions or a junction and the end of the street) like the route identifier of the route the section belongs to, the tuple identifier of this route in the routes relation, start and end position of the section on the route, section curve, and two boolean flags \var{startssmaller} and \var{dual} with the same meaning as in the routes relation. We introduce four B-Tree indexes and one R-Tree index to support faster query execution. The four B-Trees indicate the route identifier attributes in the three main relations. And the R-Tree indicates the curve attribute of the routes relation. Furthermore there are two sets which provide a fast access from one section to their adjacent sections in driving direction\footnote{We call two sections adjacent if their lanes are connected by a junction.}.

Single positions in the network are given as \dt{gpoint} values. Besides the network identifier a \dt{gpoint} consists of a route identifier, a distance from the start of the route to the position of the \dt{gpoint} and a \dt{side} value ({\var{up}, \var{down}, \var{none}). The \dt{side} value is basically necessary if the route is dual, it tells us if the position is reachable form the \var{up} or the \var{down} \dt{side} of the route. For simple routes or positions which are reachable form both sides of the route the \dt{side} value is always \var{none}.

Parts of the network, regardless if they represent paths or regions, are given as \dt{gline} values. Besides the network identifier a \dt{gline} consists of a set of \dt{route intervals}, and two boolean flags telling us if the \dt{gline} is defined and if the set of \dt{route intervals} is sorted. Each \dt{route interval} consists of a route identifier identifying the route the \dt{route interval} belongs to, and the start and the end position from the \dt{route interval} on this route. In the original paper the \dt{route interval} includes a additional parameter \dt{side} like the \dt{gpoint} does. But this parameter is not part of the implementation yet. We call the \label{sec:sortedgline} set of \dt{route intervals} sorted if the following conditions are fullfilled:
\begin{enumerate}
	\item all \dt{route intervals} are disjoint
	\item the \dt{route intervals} are stored in ascending order of their route identifiers
	\item if two disjoint \dt{route intervals} have the same route identifier the \dt{route interval} with the smaller start position is stored first
	\item for all \var{route intervals} in the set the condition: $\var{start position} \le \var{end position}$ holds
\end{enumerate}
We introduced the sorted \dt{gline} because many algorithms take profit from sorted \dt{gline} values. For example the computation if a \dt{gpoint} is inside the \dt{gline} takes O(\var{n}) for unsorted and O($\log$ \var{n}) time for sorted \dt{gline} values, if \var{n} is the number of \dt{route intervals} in the \dt{gline}.

Unfortunately not all \dt{gline} values can be stored sorted. If a \dt{gline} value represents a path between two \dt{gpoint} in the network we need the \dt{route intervals} exactly in the sequence they are used in the path. This will nearly never be a sorted set like defined before. We solved this dilemma by introducing the sorted flag. Every algorithm which can take profit from a sorted \dt{gline} values checks this flag and uses the corresponding code. We store  \dt{gline} values sorted whenever this is possible to support faster query execution.

@@

In deviation from the original network data model we extended the implementation of the \dt{mgpoint} with four additional attributes:
\begin{enumerate}
	\item The total driven distance
	\item A sorted set of \dt{route intervals} representing the positions ever traversed by the \dt{mgpoint}
	\item A boolean defined flag for the set of \dt{route intervals}
	\item A spatio-temporal minimum bounding box
\end{enumerate}
The sorted set of \dt{route intervals} was introduced, because it makes it much faster to decide if a \dt{mgpoint} ever passed a given place or not. Instead of a linear check of all \var{m} \dt{ugpoint}s of a \dt{mgpoint} we perform a binary scan on the much lower number \var{r} of \dt{route intervals}. This reduces the time complexity from O(\var{m}) to O($\log \var{r}$) for all \op{passes} operations. Logical this should be done by a sorted \dt{gline} value but the \secondo{} DBMS restricts us to use a sorted set of \dt{route intervals} instead.

The spatio-temporal minimum bounding box was introduced as parameter to the \dt{mgpoint} because the computation of this value is very expensive in the network data model. Although each unit of a \dt{mgpoint} stays on the same route at same speed the motion may follow different spatial directions, e.g. a route may lead uphill in serpentine. Not all this positions must be enclosed by a bounding box computed just from the spatial position of the \var{start} and \var{end} position of the unit. Therefore we have always to examine the spatial dimensions of the complete part of the route passed within a unit to compute the units bounding box. All spatial information of the route curve is hidden in the \dt{network} object. We have to call the route curve from the \dt{network} object to compute the spatial dimensions of the unit bounding box. If \var{r} is the number of routes of the network and \var{h} the number of half segments of the traversed part of the route curve passed in a unit we need O($\log \var{r} + \var{h}$) time to compute the bounding box for a single unit. The bounding box of the \dt{mgpoint} is the union of the bounding boxes of its \var{m} units. So the computation of a \dt{mgpoint} bounding box takes O($\var{m}(\var{h} + \log r)$) time. This is very expensive therefore the bounding box of a \dt{mgpoint} is only computed on demand or if we can get it for free. E.g. we can copy the bounding box of a \dt{mpoint} if we translate it into a \dt{mgpoint} without computational effort. But we don't maintain this attribute at every change of the \dt{mgpoint}. If the \dt{mgpoint} changes we set the bounding box attribute to be undefined and compute it again on demand if necessary.
\section{BerlinMOD and Network Data Model}
\label{sec:bmodbNetDataMod}
In the next subsections we first present our experimental setup (\ref{sec:scenario}) and the transfer from the \bmodb{} data-sets and queries in our network data model representation including a description of the indexes we build to support faster query execution (\ref{sec:Translation}) before we conclude the section with the results of our experiments in \ref{sec:results}.
\subsection{Experimental Setup}
\label{sec:scenario}
For our experiments we used a Standard PC with a Intel Pentium 4 2.93 GHz CPU, 1 GB main memory, 200 GB HDD, and Linux openSUSE 11.1 as operating system. We installed \secondo{} version 2.9.1+ and the \bmodb{} from the web.

>>>>

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
\newcommand{\secver}{3.0?}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past several data models for the representation of spatial and
spatio-temporal data objects have been developed. Among others we can categorise
the data models into data models for objects moving freely in space and data
models for moving objects that are constrained by a given network. For both
categories several different data models have been developed. Two of them, one
data model for objects moving freely in the 2D plane and one for network
constrained moving objects have been implemented in \secondo{} DBMS. Both
handle with the histories of moving objects. In this paper we present the
comparison of the capabilities of the both data models using the \bmodb{} and
propose a extension of the \bmodb{} with specialised challenges for network
constrained data models called BerlinMOD/Net. The extension enables us to
compare the capabilities of different network data models with respect to the
the specialised challenges of the network constrained data model.
\end{abstract}
\section{Introduction}
In the past several data models for the representation of spatial and
spatio-temporal data objects have been developed. Among others we can categorise
them into data models for objects moving freely in space and data models for
objects which movement is constrained by a given network. For both categories
several different data models have been presented like
\cite{335426,chenzaniolosqlst} for spatio-temporal data objects moving freely in
 space and \cite{1146465,956692,VazWolfNetMod} for spatio-temporal data objects
that are constrained by networks to name just a few.

For our experiments we choosed one example data model as a representative for
each of the both categories to compare the capabilities of the both different
data models. The category of data models for spatio-temporal objects moving
freely in space is represented by the data model presented in \cite{335426}.
And the category of network constrained data models is represented by the
network data model presented in \cite{1146465}. The advantage of this choice is,
 that both data models are implemented in the \secondo{} DBMS and use the same
 temporal representation. So we can exclude that the use of different DBMS or
temporal representations bias our results.

We used the \bmodb{} \cite{BerlinMODVLDB} to compare the capabilities of the
both data models. The \bmodb{} is best to our knowledge the first benchmark for
complete spatio-temporal database systems and it is available in the \secondo{}
 DBMS. Furthermore the \bmodb{} data model is the data model of freely movement
we use for our comparison. So we only have to transform the \bmodb{} spatial and
 spatio-temporal data types once in our network data model representation. This
avoids sources of errors and makes the correctness controll of the query results
 more easy.

In our experiments the network data model outperforms the data model of free
movement in space significantly. The total run time over all queries for the
network data model is half as long as the total run time for the data model of
free movement in the space. So the further developement of network data models
seems to be useful.

Network data models have their own challenges like network distance computing
which are not covered by the \bmodb{} yet. Therefore we propose a extension of
the \bmodb{} with a additional query set BerlinMOD/NET covering the specialised
challenges for spatio-temporal databases handling with network constrained moving
objects to enable us to compare the capabilities of different network data models
with the \bmodb{}.

We will present our work as follows: In section \ref{sec:relWork} we give a
short reminder of the underlying \secondo{} DBMS (\ref{sec:secondo}), the \bmodb{} (\ref{sec:bmodb}) and the two data models (\ref{sec:bmodbdatamod}
 and \ref{sec:netdatamod}) we used for our comparision. In section
\ref{sec:bmodbNetDataMod} we present the setup for our experiments (\ref{sec:scenario})
 and describe our transformation of the \bmodb{} data and queries in the network
 data model representation (\ref{sec:Translation}) before we show the results of
our experiments (\ref{sec:results}). In section \ref{sec:newqueries} we present
 our extension of the \bmodb{} query set. We conclude the paper with a summation
 of our results in section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
As mentioned before in the past other data models for free movement in the space \cite{335426,chenzaniolosqlst} and for network constrained movement have been
presented \cite{1146465,956692,VazWolfNetMod}. As well there are some more
benchmarks \cite{COSTBenchmark, QueriesTheodoridis} and database systems for
spatial and spatio-temporal datatypes \cite{HERMES,1054151}. But they don't
provide a combination of different implemented and supported data types together
 with a existing benchmark like the actual \secondo{} DBMS. So it is self-evident
for us to use the \secondo{} DBMS in combination with the provided data types and
the \bmodb{} to compare the capabilities of the both different data models.

In the next subsections we give short reminders of the \secondo{} database system (\ref{sec:secondo}), the \bmodb{} (\ref{sec:bmodb}), and the both data models (\ref{sec:bmodbdatamod}, \ref{sec:netdatamod}) we used in our experiments. More
detailed information for each of them can be taken from the original papers.
\subsection{Secondo}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in \cite{686903,1054151} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
 single data models. Hence \secondo{} can be easily extended by the user
implementing so called \secondo{} algebra modules to introduce  new data types
and operations on this data types. The user also may define additional viewers
for the graphical user interface or write optimization rules or cost functions
to extend the optimizer. \secondo{} is free available in the web \cite{secondoweb}
and comes with a number of already implemented spatial and spatio-temporal data types
and operations including the spatio-temporal data model of free movement in the space \ref{sec:bmodbdatamod} and the network data model \ref{sec:netdatamod}. Furthermore the \bmodb{} descripted in \ref{sec:bmodb} has been developed in the \secondo{} DBMS.
For our experiments we used the \secondo{} version \secver{}.
\subsection{BerlinMOD Benchmark}
\label{sec:bmodb}
The \bmodb{} was presented in \cite{BerlinMODVLDB} \nocite{BerlinMOD} and the
provided scripts for the data generation are implemented for the \secondo{} DBMS.
It is available in the web \cite{berlinmodweb} and provides a well defined
data-set and queries for the experimental evaluation of different moving object
data representations. The \bmodb{} emphasises the development of complete systems
and simplifies experimental repeatability pointing out the weakness and the potency
of the benchmarked systems.

The data-sets of the \bmodb{} are created using the street map of the German
captial Berlin \cite{bbike} and statistical data about the regions of Berlin \cite{bevberlin,berlinstadtatlas} as input relations.
The created moving objects represent cars driving in the streets of Berlin.
This makes it possible to use the data set of the \bmodb{} for network constrained
 data models. Every moving object has a home node and a work node and every weekday
 there will be a trip from the home node to the work node in the morning and trip
from the work node back to the home node in the afternoon. In the evening and at
the weekend randomly chosen cars spend additional trips (one in the evening and up
to six at the weekend) to different randomly chosen targets. The number of observed
cars and the duration of the observation period can be influenced by the \bmodb{}
user by setting the $SCALEFACTOR$ to different values in the data generation
script. For example at $SCALEFACTOR$ 1.0 the data generator will create 2000
moving objects observed for 28 days. Each of them sending a GPS-signal every 2
seconds. This simulated signals are simplified so that time intervals when a car
 doesn't move or moves in the same direction at the same speed are merged into
one moving step. E.g. if the car holds 8 hours in front of the work node there
 will be only one entry in the cars history of movement with a time interval of
8 hours instead of 14.400 entries one for each GPS interval.

The \bmodb{} provides two different approaches to store the histories of moving
objects. On the one hand the object-based approach (OBA) and on the other hand
the trip based approach (TBA).

In the OBA the complete history for each moving object
is kept together into one single entry. There is only one relation (dataScar)
containing one tuple for each object consisting of the spatio-temporal data of
the object (journey), the licence, the type, and the model of the object.

In the TBA we have two relations. One of them (dataMcar) contains the static data
 for each object like licence, type, and model together with an object identifier.
The other relation (dataMtrip) contains for each object identifier several tuples
each of them containing a single trip of the moving object (e.g. each time the car
drives from home node to work node) or a longer stop (e.g. the time the car holds
in front of the office).

Besides the moving objects the \bmodb{} provides sets of $QueryPoints$,
$QueryRegions$, $QueryInstants$, $QueryPeriods$, and $QueryLicences$,
 each of them containing 100 pseudo randomly generated objects (points, regions,
time instants, time intervals, and licences) used in the benchmark queries.

For this data objects the \bmodb{} provides two sets of queries. One set addresses
range queries (BerlinMOD/R) and the  other one nearest neighbour queries(BerlinMOD/NN).
 In this paper we focus on the range queries, which are the main aspect of the
\bmodb{} up to now.

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects object identity (known / unknown),
dimension (standard / spatial / temporal / spatio-temporal),
query interval (point / range / unbounded),
condition type (single object / object relations),
and aggregation (with or without aggregation).

We will present the 17 queries in more detail in section \ref{sec:queries}
together with our network data model translations.
\subsection{BerlinMOD Data Model}
\label{sec:bmodbdatamod}
The data model used by the \bmodb{} is the same data model of freely moving in
2D space presented in \cite{594784,335426,352963}. All spatial positions are
given in x,y-coordinates. A single spatial position is represented by the data
type \dt{point}. A \dt{point} consists of a pair of \dt{real} values interpreted
as x,y-coordinates in the 2D plane. Streets are represented by \dt{line} values.
 A \dt{line} value consists of a set of half segments representing the geometry
of the line in the 2D plane. Each half segment consists of two \dt{point} values
which are interpreted as start and end point of the half segment. Regions are
represented by the data type \dt{region}. A \dt{region} consists of a set of
half segments interpreted as outer (and inner) border of the region in the 2D plane.

@@

Some other data types of \secondo{} which are used in the \bmodb{} are shown in table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabular}{|l|p{9.5cm}|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type.\\
\hline
\dt{int} & Usual integer number.\\
\hline
\dt{real} & Usual real number.\\
\hline
\dt{instant} & A point in time.\\
\hline
\dt{periods} & A set of disjoint and not connected time intervals.\\
\hline
\dt{mbool} & A time dependent boolean value. The value within each \dt{ubool} will
 be constant \true{} or \false{} \\
\hline
\dt{mreal} & Time dependent real number. Each unit will be defined by a function
of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabular}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\subsection{Network Data Model}
\label{sec:netdatamod}
The central idea of the network data model presented in \cite{1146465} is that
every movement is constrained by a network and every position is given related
to this network. The data type \dt{network} is the central data type in the network
data model. All other data types of the network data model are related to a
\dt{network} by the unique network identifier that is part of each \dt{network}
object.

The \dt{network} object contains all spatial information of the represented network
in three main relations. The first relation contains the attributes of the routes
(streets) like id, route curve, route length, and two boolean flags indicating
if the route starts at the lexicographic smaller end point and if the lanes of
the route are separated like on German Highways or not. The second relation
contains all attributes of the junctions (street crossings) like the identifiers
of the first and second route crossing in the junction, the distance of the
junction from the start of the first respectively second route, tuple identifiers
 of the both routes in the routes relation, tuple identifiers of the sections
connected by this junction in the sections relation, and a connectivity code
telling us which lanes of the two routes are connected by the junction. The third
main relation of the \dt{network} object is the sections relation containing the
attributes of the sections (street parts between two junctions or a junction and
the end of the street) like the route identifier of the route the section belongs to,
 the tuple identifier of this route in the routes relation, start and end position
 of the section on the route, section curve, and two boolean flags $startssmaller$
and $dual$ with the same meaning as in the routes relation.

We introduce four B-Tree indexes and one R-Tree index to support faster query execution.
 The four B-Trees indicate the route identifier attributes in the three main relations.
 And the R-Tree indicates the curve attribute of the routes relation. Furthermore
there are two sets which provide a fast access from each section to their adjacent
sections with respect to the driving direction. Two sections are adjacent if their
lanes are connected by a junction.

Single positions in the network are given as \dt{gpoint} values. Besides the
network identifier a \dt{gpoint} consists of a route identifier, a distance from
the start of the route to the position of the \dt{gpoint} and a \dt{side} value
({$up$, $down$, $none$).

The \dt{side} value is necessary if the route is dual, it tells us if the position
is reachable form the $up$ or the $down$ \dt{side} of the route. For simple
routes or positions which are reachable form both sides of the route the \dt{side}
value is always $none$.

Parts of the network, regardless if they represent paths or regions, are given as
\dt{gline} values. Besides the network identifier a \dt{gline} consists of a set
of \dt{RouteIntervals}, and two boolean flags telling us if the \dt{gline} is defined
and if the set of \dt{RouteIntervals} is sorted.

Each \dt{RouteInterval} consists of a route identifier identifying the route
the \dt{RouteInterval} belongs to, and the start and the end position from the
\dt{RouteInterval} on this route\footnote{In the original paper the
 \dt{RouteInterval}
includes a additional parameter \dt{side} like the \dt{gpoint} does. But this
parameter is not part of the implementation yet.}. We call a \label{sec:sortedgline}
set of \dt{RouteIntervals} sorted if the following conditions are fullfilled:
\begin{enumerate}
	\item all \dt{RouteIntervals} are disjoint
	\item the \dt{RouteIntervals} are stored in ascending order of their route identifiers
	\item if two disjoint \dt{RouteIntervals} have the same route identifier the \dt{RouteInterval} with the smaller start position is stored first
	\item for all $RouteIntervals$ in the set the condition: $startPosition \le endPosition$ holds
\end{enumerate}
We introduced the sorted \dt{gline} because many algorithms take profit from sorted
\dt{gline} values. For example the computation if a \dt{gpoint} is inside a
\dt{gline} can be done in O($n$) for unsorted and O($\log n$) time for
sorted \dt{gline} values, if $n$ is the number of \dt{RouteIntervals} in the
\dt{gline}.

Unfortunately not all \dt{gline} values can be stored sorted. If a \dt{gline}
value represents a path between two \dt{gpoint} in the network, we need the
\dt{RouteIntervals} exactly in the sequence they are used in the path. This will
nearly never be a sorted set like defined before. We solved this dilemma by
introducing the sorted flag. Every algorithm which can take profit from a sorted
\dt{gline} checks this flag and uses the corresponding code. We store \dt{gline}
values sorted whenever this is possible to support faster query execution.

@@

In deviation from the original network data model we extended the implementation of the \dt{mgpoint} with four additional attributes:
\begin{enumerate}
	\item The total driven distance
	\item A sorted set of \dt{RouteIntervals} representing the positions ever
traversed by the \dt{mgpoint}
	\item A boolean defined flag for the set of \dt{RouteIntervals}
	\item A spatio-temporal minimum bounding box
\end{enumerate}
The sorted set of \dt{RouteIntervals} was introduced, because it makes it much
faster to decide if a \dt{mgpoint} ever passed a given place or not. Instead of
 a linear check of all $m$ \dt{ugpoint}s of a \dt{mgpoint} we can perform a
binary scan on the much lower number $r$ of \dt{RouteIntervals}. This reduces
the time complexity from O($m$) to O($\log r$) for all \op{passes}
operations. Logically this should be done by a sorted \dt{gline} value but the
\secondo{} DBMS restricts us to use a sorted set of \dt{RouteIntervals} instead.

The spatio-temporal minimum bounding box was introduced as parameter to the
\dt{mgpoint} because the computation of this value is very expensive in the
network data model. Although each unit of a \dt{mgpoint} stays on the same route
at same speed the motion may follow different spatial directions, e.g. a route may
lead uphill in serpentine. Not all this positions must be enclosed by a bounding box
computed just from the spatial position of the $startPosition$ and $endPosition$
of the unit. Therefore we have always to examine the spatial dimensions of the
complete part of the route passed within a unit to compute the units bounding box.
All spatial information of the route curve is hidden in the \dt{network} object.
We have to call the route curve from the \dt{network} object to compute the
spatial dimensions of the unit bounding box. If $r$ is the number of routes
of the network and $h$ the number of half segments of the traversed part of
the route curve passed in a unit we need O($\log r + h$) time to
compute the bounding box for a single unit. The bounding box of the \dt{mgpoint}
 is the union of the bounding boxes of its $m$ units. So the computation of
a \dt{mgpoint} bounding box takes O($m(h + \log r)$) time. This is a
very expensive computation and the bounding box of a \dt{mgpoint} is only computed
 on  demand or if we can get it for free. E.g. we can copy the bounding box of a
\dt{mpoint} if we translate it into a \dt{mgpoint} without computational effort.
 But we don't maintain this attribute at every change of the \dt{mgpoint}. If the
\dt{mgpoint} changes we set the bounding box attribute to be undefined and compute
 it again on demand if necessary.
\section{BerlinMOD and Network Data Model}
\label{sec:bmodbNetDataMod}
In the next subsections we first present our experimental setup (\ref{sec:scenario})
 and the transfer from the \bmodb{} data-sets and queries in our network data model  representation (\ref{sec:Translation}) including a description of the indexes
we build to support faster query execution. We conclude the section with the
results of our experiments in \ref{sec:results}.
\subsection{Experimental Setup}
\label{sec:scenario}
For our experiments we used a Standard PC with a AMD Phenom X4 Quad Core 2.95 GHz CPU, 8 GB main memory, 2 TB HDD, and Linux openSUSE 11.2 as operating system. We installed \secondo{} version \secver{} and the \bmodb{} from the web.
